The Evolution of Gene Expression in the Term Placenta of Viviparous Mammals

### 2016-04-25 poster/placenta_viviparous_poster.Rnw
add start of poster

%\documentclass[english,12pt]{article}
\documentclass[doublespacing]{bmcart_dla}
\usepackage{fontspec}
\setmainfont{FreeSans}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\setmathrm{FreeSans}
\setmathsf{FreeSans}
\setmathtt{FreeSans}
%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage[letterpaper]{geometry}
\usepackage{url}
% \usepackage{fancyhdr}
\usepackage{graphicx}
% \usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
%\usepackage[plain]{fancyref}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage{amsmath}
\usepackage{sansmath}
\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
\addbibresource{references.bib}
% \usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black,citecolor=blue]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
\usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
\renewcommand{\efloatseparator}{\mbox{}}
%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\renewcommand{\includegraphics}[2][]{}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
% \usepackage[x11names,svgnames]{xcolor}
% \usepackage{texshade}
% \usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
\usepackage{microtype}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage[capitalize]{cleveref}
% \newenvironment{narrow}[2]{%
%   \begin{list}{}{%
%       \setlength{\topsep}{0pt}%
%       \setlength{\leftmargin}{#1}%
%       \setlength{\rightmargin}{#2}%
%       \setlength{\listparindent}{\parindent}%
%       \setlength{\itemindent}{\parindent}%
%       \setlength{\parsep}{\parskip}}%
%   \item[]}{\end{list}}
% \def\newblock{\hskip}
% \newenvironment{paperquote}{%
%   \begin{quote}%
%      \it
%   }%
%   {\end{quote}}
% \renewcommand{\textfraction}{0.15}
% \renewcommand{\topfraction}{0.85}
% \renewcommand{\bottomfraction}{0.65}
% \renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
% \newenvironment{enumerate*}%
%   {\begin{enumerate}%
%     \setlength{\itemsep}{0pt}%
%     \setlength{\parskip}{0pt}}%
%   {\end{enumerate}}
% \newenvironment{itemize*}%
%   {\begin{itemize}%
%     \setlength{\itemsep}{0pt}%
%     \setlength{\parskip}{0pt}}%
%   {\end{itemize}}
% \setcounter{tocdepth}{3}
% \oddsidemargin 0.0in 
% \textwidth 6in
% \raggedbottom
% \clubpenalty = 10000
% \widowpenalty = 10000
% set a large emergency stretch to avoid issues with overfull boxes
% \setlength{\emergencystretch}{20em}
% make the afill smaller
% \renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
% \pagestyle{fancy}

\input{common_latex}

\begin{document}
\raggedright

@@

% Acronyms
\input{acronyms}

\begin{frontmatter}
\begin{fmbox}
\dochead{Research}%
\title{The evolution of gene expression in the term placenta of viviparous mammals}%
%
\author[addressref={aff1},noteref={n1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
\author[addressref={aff2},noteref={n1},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL}\fnm{Amy} \snm{Weckle}}
\author[addressref={aff3},email={jacaravas@gmail.com}]{\inits{J}\fnm{Jason} \snm{Caravas}}
\author[addressref={aff4},email={agnewd@dcpah.msu.edu}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
\author[addressref={aff5},email={kbenirschke@ucsd.edu}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
\author[addressref={aff6},email={suerumbaugh@gmail.com}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
\author[addressref={aff11},email={nevo@research.haifa.ac.il}]{\inits{E}\fnm{Eviatar} \snm{Nevo}}
\author[addressref={aff7,aff8},email={cjkim@med.wayne.edu}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
\author[addressref={aff9},email={gunter.wagner@yale.edu}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
\author[addressref={aff7},email={prbchiefstaff@med.wayne.edu}]{\inits{R}\fnm{Roberto} \snm{Romero}}
\author[addressref={aff1,aff6,aff10},corref={aff1},email={wildmand@illinois.edu}]{\inits{DE}\fnm{Derek E.} \snm{Wildman}}
\address[id=aff1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[id=aff2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
\address[id=aff3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
\address[id=aff4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
\address[id=aff5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
\address[id=aff6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
\address[id=aff7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
\address[id=aff8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
\address[id=aff9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
\address[id=aff10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
\address[id=aff11]{The Institute of Evolution, University of Haifa, Haifa, Israel}

\begin{artnotes}
\end{artnotes}

\begin{artnotes}
\note[id=n1]{Equal contributor} % note, connected to author
\end{artnotes}

\end{fmbox}
% \twocolumn

\begin{abstractbox}
\begin{abstract}
% \section{Abstract}
  \parttitle{Background} The placenta is arguably the most
  anatomically variable organ in mammals. Understanding gene
  expression patterns among placentas from multiple species
  illuminates the molecular basis of placenta function as well
  obstetrical syndromes affecting humans.

  \parttitle{Results} We measured the expression profiles of
  \Sexpr{star.logs[,length(unique(file))]} term placentas of
  \Sexpr{star.logs[,length(unique(species))]} species of mammals. We
  found lineage-specific differential expression of prolactin,
  relaxin, and galectins as well as increased expression of the
  pre-eclampsia associated genes \textit{ADAM12}, \textit{CRH}, and
  \textit{PAPPA} in humans. We also identified genes which are
  expressed in all placentas, including genes with immune-modulating
  properties (\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and
  \textit{S100A10}) and cell-cell interactions (\textit{LAMC1},
  \textit{LUM}, and \textit{LGALS1}).

  \parttitle{Conclusions} A diverse array of genes are differentially
  expressed and are likely to play a role in the anatomical
  variability of the placenta in the placentas of mammalian lineages.
  Furthermore, multiple genes which are associated with pre-eclampsia
  (\textit{CRH}, \textit{ADAM12}, \textit{KISS1}, \textit{PAPPA},
  \textit{IL1RL1}, and \textit{PLAC1}) show increased expression on
  the human lineage. Still, using the strictest filtering strategy
  we find a core set of only
  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
  genes is essential for the proper functioning of the placenta in all
  species analyzed.

\textbf{Keywords:} mammal, gene expression, evolution, RNA-seq, placenta

\end{abstract}
\end{abstractbox}
\end{frontmatter}

\section{Background}

The main role of the placenta in viviparous mammals is the transfer of
nutrients, oxygen, and waste between mother and offspring. Although
these functions are conserved among mammals, the placenta is perhaps
the most variable organ in terms of structure and form
\cite{Mossman.1937:comparativemorphogenesisofthe,Moffett.ea2006:Immunologyofplacentationin}. Species differ in
the shape of attachment (i.e., discoid, diffuse, cotyledonary,
zonary), degree of intimacy between fetal tissues and maternal tissues
(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
interdigitation of maternal-fetal tissue (i.e., villous, folded,
trabecular, lamellar, labrynthine)
\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
mammals differ drastically in the number and size of their offspring,
as well as the length of gestation, traits often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(\cref{tab:gestational_periods}).

The placenta arose at some point before the origin of extant mammals,
a group that includes marsupials, monotremes, and eutherian or
“placental mammals”. The term “placental mammals” is misleading, as
extant monotremes (such as \textit{Ornithorhynchus anatinus}, the
platypus) possess a rudimentary placenta, as do marsupials, although
the direct apposition to maternal tissues lasts for a proportionally
shorter time during gestation
\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
Marsupials also differ from eutherian mammals in that most possess
only a choriovitelline placenta, produced from the fusion of the
chorion and yolk sac, during most of gestation. Although some
eutherian mammals may possess a choriovitelline placenta at some point
in development, the major site of nutrient exchange in placentals
occurs via the chorioallantoic placenta, produced from the fusion of
the chorion and allantois
\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
the first placenta was more like that of marsupials or eutherians;
however, using phylogenetic methods, the ancestral eutherian was
reconstructed as possessing a relatively invasive hemochorial
placenta, as in humans and mouse
\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
in \cow, genes related to the thymus-specific serine protease
\textit{PRSS16} in \textit{Loxodona africana} (in this work),
 and the expansion of the growth hormone
and chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{Pan troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

One explanation for the divergent paths of the placental interface is
the distinct evolutionary interplay between maternal and fetal
strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}.
The maternal strategy is to requisition an adequate degree of
nutrients to the fetus without risking the health of the mother, while
the fetal strategy via paternal genes is to extract as many resources
from the mother as possible. Therefore, an interplay between these
extremes can set up an intricate “arms race”, involving the imprinting
of parental alleles, expansion of gene families, and enlistment of
existing genes to expression at the maternal-fetal interface
\cite{Haig1993:Geneticconflictsinhuman}.

One symptom of this maternal-fetal conflict is its potential affect on
health
\cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
addition to having an invasive hemochorial placenta, \human possess a
deeply invasive trophoblast layer that can penetrate the myometrium of
the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
Defects in placental growth are known to be associated with the
devastating condition of pre-eclampsia, which can cause hypertension,
proteinuria, and has the potential to lead to death of the mother
and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
Although cases have been documented in \textit{P. troglodytes}
\cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
\textit{Gorilla gorilla}
\cite{Thornton.ea1992:Convulsionsinpregnancyin}, pre-eclampsia has a
far higher incidence in \human pregnancies with as much as 4\% of
pregnancies being affected in the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. \textit{P.
  troglodytes} and \textit{G. gorilla} also have deeply invasive
trophoblasts, although the progression of invasion in great apes may
be slower than \human
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. However,
large-scale expression studies of great ape placentas have not been
conducted. The increasing affordability of next-generation sequencing
technology has enabled the large-scale study of specific tissues from
a wide array of species to address broader phylogenetic questions in
the evolution of gene expression
\cite{Brawand.ea2011:evolutionofgeneexpression}.

Here we use RNA-Seq
technology to sequence and quantify the mRNA environment of mammalian
placentas at term, including that of the \human, \textit{Pan paniscus}
(the sister species to \textit{P. troglodytes}), \textit{Ateles
  fusciceps}, \mouse, \galili, \carmeli, \textit{Canis familiaris},
\cow, \elephant, \textit{Dasypus novemcinctus}, and
\textit{Monodelphis domesticus}. These species represent four major
branches of eutherian mammals (Euarchontoglires, Laurasiatheria,
Afrotheria, and Xenartha) as well as marsupials. These placentas also represent a
wide array of placental morphologies as well as traits such as number
and size of offspring and length of gestation
(\cref{tab:gestational_periods}). We use this data, along with
existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual}, \textit{Equus caballus}
\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis aries}
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \textit{Sus scrofa}
\cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
differences between expression in marsupial and eutherian mammals at
term, identify the evolution of expression in important placental
genes such as hormones and galectins, identify novel lineage-specific
expression patterns, and quantify \human specific differences in gene
expression that may be related to obstetrical disorders such as
pre-eclampsia.

\begin{table}
  \begin{tabular}{c >{\raggedright \centering}p{1in} c c c}
    \toprule
    Species                       & Common Name & Gestation (days) & \multicolumn{1}{p{1in}}{\centering Average Litter Size}  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & Spider monkey & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & Cow & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & Dog & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & Nine-banded Armadillo & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
    \textit{Equus caballus}       & Horse & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & African savanna elephant & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & Human & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& Gray short-tailed opossum & 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & Mouse & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Spalax galili}   & Blind mole rat from Upper Galilee & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & Domestic sheep & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & Bonobo & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & Blind mole rat from northern Israel & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & Domestic pig (Duroc) & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
  \caption{Gestational periods in days and average liter sizes for studied species.}
  \label{tab:gestational_periods}
\end{table}

\section{Methods}

\subsection{Collection of Placental Tissue and RNA Extraction}

Fetal tissue was collected from the placenta of nine mammalian
species (\human [human], \bonobo [bonobo], \smonkey [spider monkey], \dog [domestic dog], \cow
[domestic cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \galili [blind mole rat from Upper Galilee],
\carmeli [blind mole rat from northern Israel]
\cite{Nevo2001:adaptiveradiationofblindsubterraneanmole}). Information
on specimen ID, and collector is shown for each sample in Supplemental
\cref{tab:srr_numbers}. In the case of \bonobo, \cow, and \elephant,
the placenta was collected at term upon expulsion from the mother
following delivery. Placental tissue from \human was collected upon
delivery via a cesarean section. Near-term placental tissue from \dog
was collected upon the death of the mother from a car accident.
Placental tissue from \opossum was sampled at fetal stage 32, roughly
one day prior to birth. Only the fetal portion of the placenta was
sampled at the maternal-fetal interface (i.e. villous portion for
\human, \bonobo, \dog, and \elephant; cotyledon for \cow; diffuse
placenta for \opossum). All tissues were collected and stored in
RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation could
be performed. Tissues were homogenized in Trizol, and total RNA was
isolated following the manufacturer's instructions. RNA was further
purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in conjunction
with the RNase-Free DNase Set (Qiagen, Valencia, CA), following the
manufacturer’s protocol to remove any contaminating DNA from the
sample. Total RNA was then evaluated on a Nanodrop 1000 (Thermo
Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100 (Santa
Clara, CA) to confirm concentration and quality of the isolation. RNA
quality was assessed using A260/A280 and A260/A230 absorbance values.

\subsection{Sequencing}

All RNA-Seq mRNA isolation, library construction, and Illumina
sequencing was performed by the Applied Genomics Technology Center
(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
total RNA and the standard Illumina protocol. Paired-end sequencing
was performed using the Illumina Genome Analyzer II with an insert
size of 230 base pairs (bp). Read lengths for species are shown in
Supplemental \cref{tab:ref_genome_aligned_species}.

\subsubsection{Additional Sequences}

For more complete evolutionary comparisons additional sequences of
placenta in \mouse
(n=\Sexpr{star.logs[grepl("musculus",species),length(unique(file))]})
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
(n=\Sexpr{star.logs[grepl("sapiens",species),length(unique(file))]})
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
\cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
\ac{SRA} (Supplemental \cref{tab:srr_numbers}) and analyzed using the
same methodology as RNAs sequenced specifically for this study.

\subsection{Transcriptome Assembly, Alignment, and Quantification}

The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
either the \human or \mouse transcriptome (Supplemental
\cref{tab:ref_genome_aligned_species}). For annotation using Diamond,
the match with the highest bitscore was chosen as the annotation;
transcripts where the highest scoring match had a bitscore below 40
were left unannotated.
% 
All transcriptomes were then aligned to the appropriate reference (or
assembled) genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM}
\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
%
The genome of the closely-related \textit{P. troglodytes} was used as
a reference for \bonobo. The separation between the two recognized Pan
species is generally perceived to be $\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}. %
%

@@

\subsection{Exploration of Individual Transcriptome Data}

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using topGO
\cite{topGO:2.20.0}. Gene Ontology clusters were considered
significantly enriched if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} on from the
$p$ values obtained from the Fisher Exact Test were less than or equal
to 0.05. Enrichment tests were considered with two gene lists per
transcriptome: annotated genes with expression levels of the top 5\%
of the frequency distribution of \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris});
the top 100 genes with the highest \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
Due to the paucity of annotated gene ontology terms for some species,
gene lists were compared using the \human Gene Ontology using the 1:1
\human ortholog. Enrichment was considered out of the set of all 1:1
orthologs present in each species.

\subsection{Comparative Analysis of Gene Expression}

All 1:1 orthologs present among \opossum, \textit{Pan troglodytes},
\cow, \dog, \elephant, \mouse, and \human were determined using
Ensembl v.80 gene trees via Biomart
\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the biomaRt R
package
\cite{Durinck.ea2009:Mappingidentifiersforintegration,Durinck.ea2005:BioMartandBioconductorpowerful}.
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
  sapiens" & mean_fpkm >= 100,length(unique(human_name))]} 1:1
orthologous genes with \ac{FPKM} $\ge 100$ were retained for future
analyses of gene expression evolution described below.

Significant shifts in expression levels between groups of species were
calculated using Student's t test on 1:1 orthologs using the mean of
each species to avoid large biases towards samples with $n \gg 1$ and
considered significant if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} was
$\le 0.05$. For \human and \mouse lineage specific calculations, a
type-III anova was calculated using
lme4~\cite{Bates.ea2015:FittingLinearMixed-EffectsModels} with each
species having separate random effects.

\subsection{Sequences, FPKMs, and Code}

All of the code used to produce every analysis presented in this paper
is available at
\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
sequences used are deposited in SRA, and new sequences for this work
can be obtained under the series
\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
(Supplemental \cref{tab:srr_numbers}). The mean FPKM for gene in each
species with ortholog annotations are located in the supplemental data
file all\_species\_mean\_fpkm.txt.xz and the per-sample FPKM values are in
all\_species\_per\_sample\_fpkm.txt.xz.

\section{Results}

\subsection{Transcriptome Statistics}

>>>>

\documentclass[a0paper,landscape]{tikzposter}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{fancybox}
\usepackage{booktabs}
% \usepackage{multirow}
\usepackage{acronym}
\usepackage{qrcode}
\usepackage[backend=biber,natbib=true,hyperref=true,style=nature]{biblatex}
\addbibresource{references.bib}
% \usepackage[nomargin,inline,draft]{fixme}
% \newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\renewcommand*{\bibfont}{\small}

% The textpos package is necessary to position textblocks at arbitary 
% places on the page.  Use showboxes option to show outlines of textboxes.
% \usepackage[absolute]{textpos}
\usepackage[absolute,showboxes]{textpos}

\usepackage{multirow}
\usepackage{array}

\title{The Evolution of Gene Expression in the Term Placenta of
  Viviparous Mammals}

\author{Don L. Armstrong, Michael R. McGowen, Amy Weckle, Jason
  Caravas, Dalen Agnew, Sue Savage-Rumbaugh, Eviatar Nevo, Chong J.
  Kim, Günter P. Wagner, Roberto Romero, and Derek Wildman}
\institute{Institute for Genomic Biology, Computing Genomes for
  Reproductive Health, University of Illinois, Urbana-Champaign}
\usetheme{Default}

% use this
% http://www.identitystandards.illinois.edu/graphicstandardsmanual/othermedia/posterpresentation.html
\definecolorstyle{UofI}{
  \definecolor{boldBlue}{HTML}{002058}
  \definecolor{boldOrange}{HTML}{E87722}
  \definecolor{professionalBlue}{HTML}{606EB2}
  \definecolor{professionalOrange}{HTML}{D45D00}
  }{
    \colorlet{backgroundcolor}{professionalBlue}
    \colorlet{framecolor}{black}
    \colorlet{titlefgcolor}{boldBlue}
    \colorlet{titlebgcolor}{white}
    \colorlet{blockbodybgcolor}{white}
    \colorlet{blocktitlebgcolor}{professionalOrange}
    \colorlet{blocktitlefgcolor}{boldBlue}
    }
\usecolorstyle{UofI}
\defineblockstyle{UofIMinimal}{
    titlewidthscale=1, bodywidthscale=1, titleleft,
    titleoffsetx=0pt, titleoffsety=0pt, bodyoffsetx=0pt, bodyoffsety=0pt,
    bodyverticalshift=0pt, roundedcorners=0, linewidth=0.2cm,
    titleinnersep=1cm, bodyinnersep=1cm
}{
    \begin{scope}[line width=\blocklinewidth, rounded corners=\blockroundedcorners]
       \ifBlockHasTitle %
           \draw[draw=none,fill=blockbodybgcolor]
               (blockbody.south west) rectangle (blocktitle.north east);
           \draw[color=blocktitlebgcolor, loosely dashed]
               (blocktitle.south west) -- (blocktitle.south east);%
       \else
             \draw[draw=none,fill=blockbodybgcolor]
                 (blockbody.south west) rectangle (blockbody.north east);
        \fi
    \end{scope}
}

\useblockstyle{UofIMinimal}
\definebackgroundstyle{UofIVertical}{
    \draw[line width=0pt, bottom color=backgroundcolor, 
    top color=white] (bottomleft) rectangle (topright);
}
\usebackgroundstyle{UofIVertical}
\definetitlestyle{UofITitle}{
    width=\paperwidth, roundedcorners=0, linewidth=0pt, innersep=12mm,
    titletotopverticalspace=5mm, titletoblockverticalspace=10mm,
    titlegraphictotitledistance=10pt
}{}
\usetitlestyle{UofITitle}

\makeatletter
\renewcommand\TP@maketitle{%
   \begin{minipage}{0.8\textwidth}
        \color{titlefgcolor}
        {\bfseries \Huge \sc \@title \par}
        \vspace*{1em}
        {\huge \@author \par}
        \vspace*{1em}
        {\LARGE \@institute}
    \end{minipage}%
    \hfill
    \begin{minipage}{0.2\linewidth}
      \includegraphics[width=0.8\linewidth,height=6cm,keepaspectratio]{figures/igb_wordmark}
    \end{minipage}
}
\makeatother

\begin{document}

@@

\input{acronyms}

\maketitle

\begin{columns}

  \column{0.25}

  \block{Introduction}{

%     Our approach identified MIR1914, MIR126, SNORD12, and SNORD5 as
%     uterus specific. CGB, LGALS13, and members of the PSG family were
%     identified as placenta specific. The accuracy of the classifier
%     was verified using RNAseq transcriptome data not in the training
%     set. We show the applicability of the classifier to simulated EVs
%     by down sampling reads to $10^5$ reads and biasing the reads to
%     account for preferential association of RNA motifs with EV
%     packaging machinery. When coupled with single-EV sequencing, our
%     classifier enables clinicians to monitor disease-specific markers
%     over the course of treatment without biopsies, increasing the
%     ability of clinicians to adapt treatment modalities to the
%     individual patient. 
  }

  \block{Methods}{     
}

\column{0.25}

@@

\column{0.25}

\column{0.25}

\block{Conclusions}{
}

\block{References}{

  \begin{center}
    \mbox{}\vspace{-\baselineskip}
    \printbibliography[heading=none]
  \end{center}
}

\block{Code and Datasets}{
The code and underlying data for this poster can be found at
\begin{center}
\qrcode[padding]{http://dla2.us/p/pv2016}
\url{http://dla2.us/p/pv2016}
\end{center}
% 
}
\block{}{
  \includegraphics[width=\linewidth,height=6cm,keepaspectratio]{figures/uofi_mark}
}

\end{columns}
